Species motif participation provides unique information about species risk of extinction

### 2023-04-25 manuscript/reorder_main_text.tex
Updates from Overleaf

            We also confirmed previously-known relationships between persistence and other measures of network structure or a species' participation in the network.
            To identify the relationship between large-scale network structure and mean probability of persistence within a network, and how the strength of disturbance might affect this relationship, we fit a GLM relating mean persistence to network size, connectance, disturbance, and their interactions (Eq. S9).
            As the three-way interaction was significant, we did not simplify the model.

            A consumer's probability of persistence is also known to vary with its in-degree and trophic level.
            To confirm these relationships in our simulated networks and identify how they might vary with different strengths of disturbance, we fit two GLMs with binomial error distribution relating a consumer's persistence to disturbance, in-degree or STL, and the interaction between disturbance and the measure of network participation (Eq. S10-S11).
            As with the GLMs relating persistence to motif participation and motif profiles, we initially fit GLMMs including random effects of network size and connectance and network identity but removed these terms due to model singularity.

@@

            Since a network's motif profile and a species' motif participation reflect the same network as other measures (size, connectance, degree, and trophic level), it is likely that these measures are related. 
            We tested whether the frequency of motifs in the network's motif profile were related to network size, connectance, and their interaction using a set of four GLMs with binomial error distribution (Appendix \emph{S3.1}, eq. S2).
            We tested whether the frequency of motifs in a consumer's motif participation vector were related to network size, connectance, and their interaction using a similar set of four GLMs with binomial error distribution.
            Finally, we fit two additional sets of GLMs (eight total) relating the frequency of each motif in a consumer's motif participation vector to the consumer's in-degree or trophic level (Appendix \emph{S3.1}, eq. S3, S4).

>>>>

            We also confirmed previously-known relationships between persistence and other measures of network structure or a species' participation in the network.
            To identify the relationship between large-scale network structure and mean probability of persistence within a network, and how the strength of disturbance might affect this relationship, we fit a GLM relating mean persistence to network size, connectance, disturbance, and their interactions.
            As the three-way interaction was significant, we did not simplify the model.

            A consumer's probability of persistence is also known to vary with its in-degree and trophic level.
            To confirm these relationships in our simulated networks and identify how they might vary with different strengths of disturbance, we fit two GLMs with binomial error distribution relating a consumer's persistence to disturbance, in-degree or STL, and the interaction between disturbance and the measure of network participation (\emph{Appendix S3}, Eq. S2-3).
            As with the GLMs relating persistence to motif participation and motif profiles, we initially fit GLMMs including random effects of network size and connectance and network identity but removed these terms due to model singularity.

@@

            Since a network's motif profile and a species' motif participation reflect the same network as other measures (size, connectance, degree, and trophic level), it is likely that these measures are related. 
            We tested whether the frequency of motifs in the network's motif profile were related to network size, connectance, and their interaction using a set of four GLMs with binomial error distribution (Appendix \emph{S3}, eq. S4).
            We tested whether the frequency of motifs in a consumer's motif participation vector were related to network size, connectance, and their interaction using a similar set of four GLMs with binomial error distribution.
            Finally, we fit two additional sets of GLMs (eight total) relating the frequency of each motif in a consumer's motif participation vector to the consumer's in-degree or trophic level (Appendix \emph{S3}, Eq. S5-6).

### 2023-04-25 manuscript/supplement_motifs.tex
Updates from Overleaf

    Appendices S1-S3 provide supplemental methods, while appendices S4-S8 provide supplemental results.

@@

            \begin{itemize}
            \item Construction of simulated networks
            \item Filtering of biologically unrealistic networks 
            \item Conversion of networks to acyclic form
            \item Comparison of empirical and simulated networks
            \end{itemize}

        \item Statistical analysis

            \begin{itemize}
                \item Persistence and other network properties (Eqn. S2-3)
                \item Motif participation, profiles and network properties (Eqn. S4-6)
                \item Network motif profiles and global network structure (Eqn. S7)
            \end{itemize}

        \item Persistence and motif participation
            \begin{itemize}
                \item Contains Table S1
            \end{itemize}

        \item Participation and network properties

            \begin{itemize}
                \item Contains Table S2
                \item Contains Fig. S1
            \end{itemize}

        \item Persistence, degree, and trophic level

            \begin{itemize}
                \item Contains Table S3
                \item Contains Fig. S2-3
            \end{itemize}

@@

        \item Persistence and network motif profiles

            \begin{itemize}
                \item Contains Table S4
            \end{itemize}

        \item Motif profiles and global properties

            \begin{itemize}
                \item Contains Figure S5
                \item Contains Table S5
            \end{itemize}

        \item Persistence and global network properties
            \begin{itemize}
                \item Contains Figures S6, S7
                \item Contains Table S6
            \end{itemize}

@@

\section{Statistical analysis}

@@

    \textbf{Network motif profiles, size, and connectance}

        As well as affecting persistence directly, global network structure (i.e., size and connectance) may affect the meso-scale structure of the network.
        This is especially likely since we include only networks which can retain all initial species after 1000 rounds of simulated Lotka-Volterra population dynamics~\citep{Cirtwill2021_inprep}:  some meso-scale structures might be more likely to allow all species to persist than others in networks with different size and connectance constraints.
        To test for this possibility, we fit four general linear models with binomial error distribution (one per motif) relating the proportion of a motif in a network's motif profile to network size, connectance, and their interaction:
        \begin{equation}
            \Bar{\rho_{n}} \approx \Sigma_{n} + \zeta_{n} + \Sigma_{n}\zeta_{n}
        \end{equation}
        where all symbols are as in equation~\ref{partic_SC}.

@@

\section{Persistence and motif participation} [updated]

    Persistence was significantly related to the proportion of each motif in the species' role (Table~\ref{tab:proportion}). Persistence tended to increase with increasing proportions of the direct competition and, at high levels of disturbance, increasing frequencies of the apparent competition motif. Persistence tended to decrease with increasing proportions of the omnivory motif and, at high levels of disturbance, the three-species chain.

    % Updated with new random effect, converted to data-scale
    \begin{table}[ht!]
        \centering
        \caption{Coefficients ($\beta$) for general linear models relating persistence to the proportion of a species' role made up by the focal motif, the level of disturbance to basal resources (corresponding to a probability of extinction ranging between 0.1 and 0.5), and their interaction  (Equation~\ref{propreq}). Coefficients in \textbf{bold} are significant ($\alpha$\textless0.005). Models were fit using a binomial error distribution and logit link function. Predictors were centered and scaled before model fitting. We also provide McFadden's pseudo-$R^2$ for each model.}
        \label{tab:proportion}                \footnotesize
        \begin{tabular}{c|c c c c | c }
        Motif & Intercept & Proportion & Disturbance & Interaction & pseudo-$R^2$  \\
        \hline
        3-sp Chain & \textbf{2.90} & \textbf{-0.204} & \textbf{-5.42} & \textbf{0.138} & 0.895 \\
        App. Com. & \textbf{0.317} & -2.63$\times10^{-3}$ & \textbf{-5.98478} & \textbf{1.95$\times10^{-2}$} & 0.895 \\
        Dir. Com. & \textbf{0.317} & \textbf{3.03$\times10^{-2}$} & \textbf{-0.599} & -1.67$\times10^{-3}$ &  0.897 \\
        Omnivory & \textbf{0.317} & \textbf{-1.81$\times10^{-2}$} & \textbf{-0.598} & \textbf{-7.76$\times10^{-3}$} & 0.895 \\
        \end{tabular}
    \end{table}        
\clearpage

\section{Participation and network properties} [updated]

@@

    \begin{figure}[ht!]
        \centering
        \includegraphics[width=.9\textwidth]{figures/participation_vs_SC.eps}
        \caption{The proportion of each motif in consumer's participation vectors varied with connectance, network size, and their interaction (except for apparent competition). We show relationships for the smallest and largest values of network size (S) and connectance (C).}
        \label{fig:roles_vs_SC}
    \end{figure}

@@

\section{Persistence, degree, and trophic level} [updated]

    Persistence varies with in-degree, trophic level, and their interactions with disturbance.

    % Updated with network-level random effect
    \begin{table}[hb!]
        \caption{Coeffcients ($\beta$) for general linear mixed-effect models of consumers' probabilities of persistence against simple measures of species' role (STL or in-degree), level of disturbance to basal resources, and their interaction, as well as a random effect of global network properties (size and connectance). The model for STL was singular when including the random effect, so we removed this term and re-fit the model as a general linear model with fixed effects only. All models had binomial error distribution. Predictors were centered and scaled before model fitting. All $p$-values were \textless0.001. The level of disturbance is reflected by basal resources having a baseline probability of extinction ranging between $0.1 - 0.5$.  We also provide the marginal $R^2_M$ (fixed effects only) and conditional $R^2_C$ (fixed and random effects) for the in-degree model and the pseudo-$R^2$ for the STL model.}
        \label{tab:per_vs_TLdeg}
        \centering
        \begin{tabular}{l|c  c  c | c c  }
            Simple role measure & Role & Disturbance & Interaction & R$^2_M$ & R$^2_C$ \\
            \hline
            STL & -0.0977 & -0.600 & 0.0302 & 0.920 & NA \\
            in-degree & -1.30$\times10^{-3}$ & -0.139 & -9.08$\times10^{-3}$ & 0.903 & 0.914 \\
        \end{tabular}
    \end{table}

@@

\section{Persistence and whole-network motif profiles} [[updated]]

    % Updated to use mean instead of all species
    \begin{table}[hb!]
        \caption{Coefficients ($\beta$) for models relating persistence to the proportion of a network profile made up by a focal motif, the level of disturbance to basal resources (corresponding to a probability of extinction ranging between 0.1 and 0.5), and their interaction. Coefficients in \textbf{bold} are significant ($\alpha$\textless0.005). Predictors were centered and scaled before model fitting; coefficients refer to a unit increase in predictors. We also provide McFadden's pseudo-$R^2$ value for each regression. }
        \label{motif_profile_tab}
        \centering
        \footnotesize
        \begin{tabular}{c|c c c c c | c }
        Motif & Intercept & Proportion & Disturbance & Interaction &  pseudo-$R^2$ \\
            \hline
            Three-species chain & \textbf{0.326} &  0.0260 & \textbf{-0.595} & -2.53$\times10^{-3}$ & 0.958 \\
            Apparent competition & \textbf{0.326} & \textbf{0.0501} & \textbf{-0.596} & -2.00$\times10^{-3}$ & 0.962 \\
            Direct competition & \textbf{0.326} & 0.0151 & \textbf{-0.595} & -5.34$\times10^{-4}$ & 0.956 \\
            Omnivory & \textbf{0.326} & \textbf{-0.0626} & \textbf{-0.596} & 3.22$\times10^{-3}$ & 0.966 \\
        \end{tabular}
    \end{table}
\clearpage

\section{Motif profiles and global properties} [[updated]]
    % No update needed

@@

   \begin{figure}[hb!]
       \centering
       \includegraphics[width=.75\textwidth]{figures/proportion_profile_dispersion.eps}
       \caption{Dispersion of normalized motif profiles (i.e., proportions of each motif) within a network about the centroid for that combination of network size and connectance. Each circle represents a single network; circle colour indicates network size. Scatter has been introduced about each level of connectance to increase visual clarity; the levels of connectance we consider are indicated by vertical black lines.}
       \label{dispersion_normmotifs}
    \end{figure}

@@

\section{Persistence and global network properties} [updated]

    Connectance had a smaller, but significant, effect on consumer persistence than disturbance (Figs.~\ref{fig:lm_CS}, Table~\ref{tab:per_vs_SC}).
    Network size had no effect on consumer persistence in these simulated networks.

    % updated to use network means rather than all species
    \begin{figure}[hb!]
        \centering
        \includegraphics[width=0.9\textwidth]{figures/persistence_vs_SC_lm.eps}
        \caption{Persistence decreased strongly with increasing probability of disturbance to basal resources and decreased slightly, but significantly, with increasing connectance. The effects of network size and interactions between predictors were not significant. }
        \label{fig:lm_CS}
    \end{figure}

    % \begin{figure}[hb!]
    %     \centering
    %   \includegraphics[width=\textwidth]{figures/heatmap_CS_BPcompare.eps}
    %    \caption{Heat maps of average consumer persistence in networks with varying connectance (y-axis) and network size (x-axis) at two levels of disturbance to basal resources. Persistence decreases with darker color in the heat map. When basal species are not disturbed, all species experience a baseline extinction risk of $\pi = 0.1$. When basal species are highly disturbed, their baseline extinction risk rises to $\pi = 0.5$ while the baseline risk of consumers is not changed.}
    %    \label{fig:heatmap_CS}
    % \end{figure}

    % Updated to use mean rather than all species.
    \begin{table}[hb!]
        \caption{Mean persistence of consumers within a network depends on connectance (C) and the level of disturbance to basal resources (D), but not network size (S) or interactions between properties. We show effect sizes ($\beta$) and $p$-values for each coefficient in a general linear model (fit using a binomial error distribution and logic link function). All predictors were centered and scaled before fitting the model. The level of disturbance corresponds to basal resources having a probability of extinction ranging between $0.1 - 0.5$ (see \emph{Methods}). McFadden's pseudo-$R^2$=0.966.
        }
        \label{tab:per_vs_SC}
        \centering
        \begin{tabular}{c|c c |}
            Predictor & $\beta$ & $p$-value \\
            \hline
            Intercept & \textbf{0.326} & \textless0.001 \\
            Size & -7.21$\times10^{-4}$  & 0.964 \\
            Connectance & \textbf{-6.12$\times10^{-2}$}& \textless0.001 \\
            Disturbance & \textbf{-0.596} & \textless0.001 \\
            S:C & -4.48$\times10^{-3}$ & 0.777 \\
            S:D & -1.63$\times10^{-2}$ & 0.316 \\
            C:D & -4.52$\times10^{-3}$ & 0.781 \\
            S:C:D & -1.35$\times10^{-3}$ & 0.934 \\
        \end{tabular}
    \end{table}

>>>>

    Appendices S1-S3 provide supplemental methods, while appendices S4-S6 provide supplemental results.

@@

            \begin{itemize}
            \item Construction of simulated networks
            \item Filtering of biologically unrealistic networks 
            \item Conversion of networks to acyclic form
            \item Comparison of empirical and simulated networks
            \item Contains Fig. S1
            \end{itemize}

        \item Supplemental statistical analysis

            \begin{itemize}
                \item Persistence and other network properties (Eqn. S2-3)
                \item Motif participation, profiles and network properties (Eqn. S4-6)
            \end{itemize}

        \item Results: species motif participation 
            \begin{itemize}
                \item Supplemental results relating species motif participation to persistence and to other network properties (S, C, in-degree, STL)
                \item Contains Table S1-2
                \item Contains Fig. S2
            \end{itemize}

        \item Results: persistence and other network properties

            \begin{itemize}
                \item Identifying relationships between persistence and network size, connectance, in-degree, and STL to provide context for motif relationships.
                \item Contains Tables S3-4
                \item Contains Fig. S3
            \end{itemize}

@@

        \item Results: network motif profiles

            \begin{itemize}
                \item Relating motif profiles to persistence (see main text for equation, figure) and to network size and connectance
                \item Contains Tables S5-6
            \end{itemize}

@@

\section{Supplemental statistical analysis}

@@

@@

\section{Results: motif participation}

    \subsection{Participation and persistence}

    % Updated with new random effect, converted to data-scale
    \begin{table}[ht!]
        \centering
        \caption{Coefficients ($\beta$) for general linear models relating persistence to the proportion of a species' role made up by the focal motif, the level of disturbance to basal resources (corresponding to a probability of extinction ranging between 0.1 and 0.5), and their interaction  (Eq. 1, main text). Coefficients in \textbf{bold} are significant ($\alpha$\textless0.005). Models were fit using a binomial error distribution and logit link function. Predictors were centered and scaled before model fitting. We also provide McFadden's pseudo-$R^2$ for each model.}
        \label{tab:proportion}                \footnotesize
        \begin{tabular}{c|c c c c | c }
        Motif & Intercept & Proportion & Disturbance & Interaction & pseudo-$R^2$  \\
        \hline
        3-sp Chain & \textbf{2.90} & \textbf{-0.204} & \textbf{-5.42} & \textbf{0.138} & 0.895 \\
        App. Com. & \textbf{0.317} & -2.63$\times10^{-3}$ & \textbf{-5.98478} & \textbf{1.95$\times10^{-2}$} & 0.895 \\
        Dir. Com. & \textbf{0.317} & \textbf{3.03$\times10^{-2}$} & \textbf{-0.599} & -1.67$\times10^{-3}$ &  0.897 \\
        Omnivory & \textbf{0.317} & \textbf{-1.81$\times10^{-2}$} & \textbf{-0.598} & \textbf{-7.76$\times10^{-3}$} & 0.895 \\
        \end{tabular}
    \end{table}

\subsection{Participation and network properties}

@@

    \begin{figure}[hb!]
        \centering
        \includegraphics[width=.9\textwidth]{figures/participation_vs_SC.eps}
        \caption{The proportion of each motif in consumer's participation vectors varied with connectance, network size, and their interaction (except for apparent competition). We show relationships for the smallest and largest values of network size (S) and connectance (C).}
        \label{fig:roles_vs_SC}
    \end{figure}

@@

\section{Results: persistence and other network properties}

    \begin{figure}[ht!]
        \centering
        \includegraphics[width=0.9\textwidth]{figures/persistence_vs_SC_lm.eps}
        \caption{Persistence decreased strongly with increasing probability of disturbance to basal resources and decreased slightly, but significantly, with increasing connectance.}
        \label{fig:lm_CS}
    \end{figure}

    % Updated to use mean rather than all species.
    \begin{table}[hb!]
        \caption{Mean persistence of consumers within a network depends on connectance (C) and the level of disturbance to basal resources (D), but not network size (S) or interactions between properties. We show effect sizes ($\beta$) and $p$-values for each coefficient, except the intercept (see Appendix S3 for details). McFadden's pseudo-$R^2$=0.966. }
        \label{tab:per_vs_SC}
        \centering
        \footnotesize
        \begin{tabular}{c| c c c c c c c}
              & S & C & D & S:C & S:D & C:D & S:C:D \\
            \hline
             $\beta$ &	 -7.21$\times10^{-4}$  	& \textbf{-6.12$\times10^{-2}$}	& \textbf{-0.596} &	 -4.48$\times10^{-3}$ &	 -1.63$\times10^{-2}$ 	& -4.52$\times10^{-3}$ &	 -1.35$\times10^{-3}$ \\
             $p$-value &		 0.964 &	 \textless0.001 &	 \textless0.001 &	 0.777 &	 0.316 &	 0.781 &	 0.934 \\
                    \end{tabular}
                \end{table}

    % Updated with network-level random effect
    \begin{table}[hb!]
        \caption{Coeffcients ($\beta$) for general linear mixed-effect models of consumers' probabilities of persistence against simple measures of species' role (STL or in-degree), level of disturbance to basal resources, and their interaction, as well as a random effect of global network properties (size and connectance). The model for STL was singular when including the random effect, so we removed this term and re-fit the model as a general linear model with fixed effects only. All models had binomial error distribution. Predictors were centered and scaled before model fitting. All $p$-values were \textless0.001. The level of disturbance is reflected by basal resources having a baseline probability of extinction ranging between $0.1 - 0.5$.  We also provide the marginal $R^2_M$ (fixed effects only) and conditional $R^2_C$ (fixed and random effects) for the in-degree model and the pseudo-$R^2$ for the STL model.}
        \label{tab:per_vs_TLdeg}
        \centering
        \begin{tabular}{l|c  c  c | c c  }
            Simple role measure & Role & Disturbance & Interaction & R$^2_M$ & R$^2_C$ \\
            \hline
            STL & \textbf{-0.0977} & \textbf{-0.600} & \textbf{0.0302} & \textbf{0.920} & NA \\
            in-degree & \textbf{-1.30$\times10^{-3}$} & \textbf{-0.139} & \textbf{-9.08$\times10^{-3}$} & 0.903 & 0.914 \\
        \end{tabular}
    \end{table}

@@

\section{Supplemental results: network motif profiles}

    \subsection{Persistence and whole-network motif profiles}

    % Updated to use mean instead of all species
    \begin{table}[ht!]
        \caption{These results correspond to equation 2, main text. Coefficients ($\beta$) for models relating persistence to the proportion of a network profile made up by a focal motif, the level of disturbance to basal resources (corresponding to a probability of extinction ranging between 0.1 and 0.5), and their interaction. Coefficients in \textbf{bold} are significant ($\alpha$\textless0.005). Predictors were centered and scaled before model fitting; coefficients refer to a unit increase in predictors. We also provide McFadden's pseudo-$R^2$ value for each regression. }
        \label{motif_profile_tab}
        \centering
        \footnotesize
        \begin{tabular}{c|c c c c c | c }
        Motif & Intercept & Proportion & Disturbance & Interaction &  pseudo-$R^2$ \\
            \hline
            Three-species chain & \textbf{0.326} &  0.0260 & \textbf{-0.595} & -2.53$\times10^{-3}$ & 0.958 \\
            Apparent competition & \textbf{0.326} & \textbf{0.0501} & \textbf{-0.596} & -2.00$\times10^{-3}$ & 0.962 \\
            Direct competition & \textbf{0.326} & 0.0151 & \textbf{-0.595} & -5.34$\times10^{-4}$ & 0.956 \\
            Omnivory & \textbf{0.326} & \textbf{-0.0626} & \textbf{-0.596} & 3.22$\times10^{-3}$ & 0.966 \\
        \end{tabular}
    \end{table}

    \subsection{Motif profiles and network size, connectance}
    % No update needed

@@

       % \begin{figure}[hb!]
       %     \centering
       %     \includegraphics[width=.75\textwidth]{figures/proportion_profile_dispersion.eps}
       %     \caption{Dispersion of normalized motif profiles (i.e., proportions of each motif) within a network about the centroid for that combination of network size and connectance. Each circle represents a single network; circle colour indicates network size. Scatter has been introduced about each level of connectance to increase visual clarity; the levels of connectance we consider are indicated by vertical black lines.}
       %     \label{dispersion_normmotifs}
       %  \end{figure}

@@
